---
title: "Tutorial for the WGCNA package for R:
  I. Network analysis of liver expression data in female mice
    2.a Automatic network construction and module detection"
output:
  html_document:
    keep_md: yes
    df_print: paged
    toc: yes
    toc_depth: '2'
  pdf_document:
    fig_caption: yes
---

# Preliminaries: setting up the R session
Here we assume that a new R session has just been started. We load the WGCNA package, set up basic parameters and load data saved in the first part of the tutorial.
**Important note**: The code below uses parallel computation where multiple cores are available. This works well when R is run from a terminal or from the Graphical User Interface (GUI) shipped with R itself, but at present it does not work with RStudio and possibly other third-party R environments. If you use RStudio or other third-party R environments, skip the `enableWGCNAThreads()` call below.
```{r}
# Display the current working directory
getwd()
# If necessary, change the path below to the directory where the data files are stored.
# "." means current directory. On Windows use a forward slash / instead of the usual \.
workingDir <- "."
setwd(workingDir)
# Load the WGCNA package
library(WGCNA)
# The following setting is important, do not omit.
options(stringsAsFactors = FALSE)
# Allow multi-threading within WGCNA. This helps speed up certain calculations.
# At present this call is necessary for the code to work.
# Any error here may be ignored but you may want to update WGCNA if you see one.
# Caution: skip this line if you run RStudio or other third-party R environments.
# See note above.
enableWGCNAThreads()
# Load the data saved in the first part
lnames <- load(file = "../data/FemaleLiver-01-dataInput.RData")
# The variable lnames contains the names of loaded variables.
lnames
```
We have loaded the variables datExpr and datTraits containing the expression and trait data, respectively.

# Automatic construction of the gene network and identification of modules
This step is the bedrock of all network analyses using the WGCNA methodology. We present three different ways of
constructing a network and identifying modules:
  1. Using a convenient 1-step network construction and module detection function, suitable for users wishing to arrive
at the result with minimum effort;
  2. Step-by-step network construction and module detection for users who would like to experiment with customized/alternate methods;
  3. An automatic block-wise network construction and module detection method for users who wish to analyze data
sets too large to be analyzed all in one.
In this tutorial section, we illustrate the 1-step, automatic network construction and module detection.

## Automatic network construction and module detection
### Choosing the soft-thresholding power: analysis of network topology
Constructing a weighted gene network entails the choice of the soft thresholding power *β* to which co-expression similarity is raised to calculate adjacency [1]. The authors of [1] have proposed to choose the soft thresholding power based on the criterion of approximate scale-free topology. We refer the reader to that work for more details; here we illustrate the use of the function `pickSoftThreshold` that performs the analysis of network topology and aids the user in choosing a proper soft-thresholding power. The user chooses a set of candidate powers (the function provides suitable default values), and the function returns a set of network indices that should be inspected, for example as follows:
```{r fig.cap="\\label{fig:figs}Analysis of network topology for various soft-thresholding powers. The left panel shows the scale-free fit index (*y*-axis) as a function of the soft-thresholding power (*x*-axis). The right panel displays the mean connectivity (degree, *y*-axis) as a function of the soft-thresholding power (*x*-axis)."}
# Choose a set of soft-thresholding powers
powers <- c(c(1:10), seq(from = 12, to = 20, by = 2))
# Call the network topology analysis function
sft <- pickSoftThreshold(datExpr, powerVector = powers, verbose = 5)
# Plot the results:
par(mfrow = c(1, 2))
cex1 <- 0.9
# Scale-free topology fit index as a function of the soft-thresholding power
plot(sft$fitIndices[, 1], -sign(sft$fitIndices[, 3]) * sft$fitIndices[, 2],
  xlab = "Soft Threshold (power)", ylab = "Scale Free Topology Model Fit,signed R^2", type = "n",
  main = paste("Scale independence")
)
text(sft$fitIndices[, 1], -sign(sft$fitIndices[, 3]) * sft$fitIndices[, 2],
  labels = powers, cex = cex1, col = "red"
)
# this line corresponds to using an R^2 cut-off of h
abline(h = 0.90, col = "red")
# Mean connectivity as a function of the soft-thresholding power
plot(sft$fitIndices[, 1], sft$fitIndices[, 5],
  xlab = "Soft Threshold (power)", ylab = "Mean Connectivity", type = "n",
  main = paste("Mean connectivity")
)
text(sft$fitIndices[, 1], sft$fitIndices[, 5], labels = powers, cex = cex1, col = "red")
```
The result is shown in Figure \ref{fig:figs}. We choose the power 6, which is the lowest power for which the scale-free topology fit
index curve flattens out upon reaching a high value (in this case, roughly 0.90).

### One-step network construction and module detection
Constructing the gene network and identifying modules is now a simple function call:
```{r}
net <- blockwiseModules(datExpr,
  power = 6,
  TOMType = "unsigned",
  minModuleSize = 30,
  reassignThreshold = 0, 
  mergeCutHeight = 0.25,
  numericLabels = TRUE, 
  pamRespectsDendro = FALSE,
  saveTOMs = TRUE,
  saveTOMFileBase = "../data/femaleMouseTOM",
  verbose = 3
)
```

We have chosen the soft thresholding power 6, a relatively large minimum module size of 30, and a medium sensitivity (deepSplit=2) to cluster splitting. The parameter `mergeCutHeight` is the threshold for merging of modules. We have also instructed the function to return numeric, rather than color, labels for modules, and to save the Topological Overlap Matrix. The output of the function may seem somewhat cryptic, but it is easy to use. For example, `net$colors` contains the module assignment, and `net$MEs` contains the module eigengenes of the modules. 

**A word of caution** for the readers who would like to adapt this code for their own data. The function `blockwiseModules` has many parameters, and in this example most of them are left at their default value. We have attempted to provide reasonable default values, but they may not be appropriate for the particular data set the reader wishes to analyze. We encourage the user to read the help file provided within the package in the R environment and experiment with tweaking the network construction and module detection parameters. The potential reward is, of course, better (biologically more relevant) results of the analysis. 

**A second word of caution** concerning block size. In particular, the parameter `maxBlockSize` tells the function how large the largest block can be that the reader’s computer can handle. The default value is 5000, which is appropriate for most modern desktops. Note that if this code were to be used to analyze a data set with more than 5000 probes, the function `blockwiseModules` will split the data set into several blocks. This will *break some of the plotting code below*, that is executing the code will lead to errors. Readers wishing to analyze larger data sets need to do one of the following:

• If the reader has access to a large workstation with more than 4 GB of memory, the parameter `maxBlockSize`can be increased. A 16GB workstation should handle up to 20000 probes; a 32GB workstation should handle perhaps 30000. A 4GB standard desktop or a laptop may handle up to 8000-10000 probes, depending on operating system and ihow much memory is in use by other running programs.

• If a computer with large-enough memory is not available, the reader should follow Section 2.c, Dealing with large datasets, and adapt the code presented there for their needs. In general it is preferable to analyze a data set in one block if possible, although in Section 2.c we present a comparison of block-wise and single-block analysis that indicates that the results are very similar. We now return to the network analysis. To see how many modules were identified and what the module sizes are, one can use table(net$colors). Its output is
```{r}
table(net$colors)
```
and indicates that there are 18 modules, labeled 1 through 18 in order of descending size, with sizes ranging from 609 to 34 genes. The label 0 is reserved for genes outside of all modules. The hierarchical clustering dendrogram (tree) used for the module identification is returned in `net$dendrograms[[1]]`. The dendrogram can be displayed together with the color assignment using the following code:
```{r fig.cap="\\label{fig:figs}Clustering dendrogram of genes, with dissimilarity based on topological overlap, together with assigned module colors. "}
# Convert labels to colors for plotting
mergedColors <- labels2colors(net$colors)
# Plot the dendrogram and the module colors underneath
plotDendroAndColors(net$dendrograms[[1]], mergedColors[net$blockGenes[[1]]],
  "Module colors",
  dendroLabels = FALSE, hang = 0.03,
  addGuide = TRUE, guideHang = 0.05
)
```

The resulting plot is shown in Figure \ref{fig:figs}. We note that if the user would like to change some of the tree cut, module membership, and module merging criteria, the package provides the function `recutBlockwiseTrees` that can apply modified criteria without having to recompute the network and the clustering dendrogram. This may save a substantial amount of time. We now save the module assignment and module eigengene information necessary for subsequent analysis.
```{r}
moduleLabels <- net$colors
moduleColors <- labels2colors(net$colors)
MEs <- net$MEs
geneTree <- net$dendrograms[[1]]
save(MEs, moduleLabels, moduleColors, geneTree,
  file = "../data/FemaleLiver-02-networkConstruction-auto.RData"
)
```

References
[1] B. Zhang and S. Horvath. A general framework for weighted gene co-expression network analysis. Statistical